---
title: "606 Project Proposal"
author: "Mei Qi Ng"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

### Data Preparation

```{r load-data, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
# load data
library(tidyverse)

student_data <- read.csv("student-mat.csv", sep = ";")

head(student_data)
```

### Research question

Is the amount of free time or the number of past class failures associated with a student's final grade? 

### Hypotheses

**Hypothesis #1:** Free time vs Grade

-   **Null Hypothesis (**$H_0$): A student's amount of free time is not associated with their final math grade. The average final grade is the same across all levels of free time.
-   **Alternative Hypothesis (**$H_A$): A student's amount of free time is associated with their final math grade. The average final grade is not the same across all levels of free time.

**Hypothesis #2:** Past Failures vs. Grades

-   **Null Hypothesis (**$H_0$): The number of past class failures is not a significant predictor of a student's final math grade.
-   **Alternative Hypothesis (**$H_A$): The number of past class failures is a significant predictor of a student's final math grade.

### Cases

Each case is a student in secondary school with a total of 395 cases in this data set.

### Data collection

**This data was collected during the 2005-2006 school year from 2 public school in the Alentejo region of Portugal (Gabriel Pereira and Monsinho da Silveira). The researchers collected the data by combining official school reports (grades and absences) with paper questionnaires given out to students during classes to gather demographic and lifestyle information**

### Type of study

**This study is an observation study. The researchers did not randomly assign treatments or change the student's environment. The student's behaviors, background, and academic results were all recorded through survey and school records**

### Describe your variables?

**Dependent Variable (Response):** G3 (Final math grade). This is a quantitative (numerical) variable, measured on a discrete scale from 0 to 20.

**Independent Variable 1 (Explanatory):** freetime (Free time after school). This is a qualitative variable, measured on a self-reported scale from 1 (very low) to 5 (very high).

**Independent Variable 2 (Explanatory):** failures (Number of past class failures). This is a quantitative variable, showing a count (0, 1, 2, or 3+).

**If you are are running a regression or similar model, which one is your dependent variable?**

-   Dependent variable is G3 (The student's final math grade - quantitative)

### Data Source

**Citation: Cortez, Paulo. (2014). Student Performance. UCI Machine Learning Repository. <https://doi.org/10.24432/C5TG7T>.**

### Relevant summary statistics

**Provide summary statistics for each the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r}
# Summary of dependent variable (Final Grade)
summary(student_data$G3)

# Summary of the independent variables
summary(student_data$freetime)
table(student_data$failures) # table() to see the count of each failure category

# Visualization 1: Final Grade by Free Time
ggplot(student_data, aes(x= factor(freetime), y = G3))+
  geom_boxplot(fill = "lightblue", color = "blue")+
  labs(title = "Relationship between Free Time and Final Math Grade",
        x = "Amount of Free Time (1 = Very Low, 5 = Very High)",
        y  = "Final Grade (0 - 20)")

# Visualization 2: Final Grade by Past Failures
ggplot(student_data, aes(x= factor(failures), y = G3))+
  geom_boxplot(fill = "red", color = "coral")+
  labs(title = "Relationship between Past Failure and Final Math Grade",
        x = "Number of Past Class Failures",
        y  = "Final Grade (0 - 20)")
```

**Analysis of Free Time vs. Final Grade: The visualization reveals a non-linear relationship between a student's self-reported free time and their final math grade. Students reporting a "2" (Low) for free time exhibit the highest median grade. This suggests a potential "sweet spot" of structured time; these students may be heavily engaged in organized extracurricular activities that reduce their free time but build strong discipline and time-management skills. Conversely, students at the lowest extreme (1) may be burdened by exhausting external obligations like jobs, while those at the highest extremes (4 and 5) may lack the structured routine necessary for peak academic achievement. Furthermore, because this data is self-reported, highly driven students with rigorous study habits may simply perceive themselves as having less free time.**


**Analysis of Past Failures vs. Final Grade: The visualization demonstrates a clear, negative relationship between a student's number of past class failures and their final math grade. As the number of past failures increases from 0 to 3, the median final grade slowly declines. Students with zero previous failures exhibit the highest median grade, which is expected as it suggests a solid academic foundation. Conversely, the consistent drop in grades for students with 1, 2, or 3 past failures indicates that these students are likely facing compounding challenges. This downward trend suggests that past failures are a strong indicator of current academic struggles, potentially pointing to unaddressed gaps in base math knowledge or external obstacles that continue to interfere with their education.**

### Assumption
```{r}
# Check: Distribution of Final Grades
ggplot(student_data, aes(x = G3)) +
  geom_histogram(binwidth = 1, fill = "green", color = "black") +
  labs(title = "Distribution of Final Math Grades (G3)",
       x = "Final Grade (0 - 20)",
       y = "Number of Students")
```
Other than the outlier at zero , the rest of the data (Grade 4 - 19) forms a normal bell curved shaped distribution; peaking at 10 and taper off at the sides. The assumption of normality is mostly met. 

Zero Anomaly: Around 39 students scored 0. This is very likely due to students who droppped the class/did not show up for the final exam as we are looking at the Portuguese school system. 
Spike at Zero = Limitation of the Analysis **

ANOVA Tests 
```{r}
# Run Test 1: ANOVA for Free Time
anova_freetime <- aov(G3 ~ factor(freetime), data = student_data)  #use factor to force R to treat as separate groups

# Summarize Hypothesis 1
print("ANOVA Summary: Free Time vs Final Grade")
summary(anova_freetime)
```


```{r}
# Run Test 2: ANOVA for Past Failures
anova_failures <- aov(G3 ~ factor(failures), data = student_data)
# Summarize Hypothesis 2
print("ANOVA Summary: Past Failures vs Final Grade")
summary(anova_failures)

```
Test 1: Free Time vs. Final Grade
P-Value: On Pr(>F) column --> p-value is 0.0657.

Interpretation: Generally, the standard threshold for significance is 0.05. Because 0.0657 is greater than 0.05, this result is not statistically significant.

The Conclusion: We fail to reject the Null Hypothesis. Even though we saw that interesting "sweet spot" at level 2 in the boxplot, this ANOVA tells us that we cannot confidently prove that free time has a significant relationship with the final grade. The variations we saw might just be due to random chance within this specific sample of 395 students.

Test 2: Past Failures vs. Final Grade
The P-Value: Your p-value is 1.64e-12.

The Interpretation: The e-12 is scientific notation. It means you need to move the decimal point 12 spaces to the left. That number is actually 0.00000000000164. Because this is drastically less than 0.05, this result is highly statistically significant (which is also why R put three little *** stars next to it!).

The Conclusion: We reject the Null Hypothesis. We have strong statistical evidence that the number of past failures is significantly associated with a student's final math grade. That downward staircase trend we saw in the red boxplot is very real.
