---
title: 'Lab 2: Introduction to data'
author: "Mei Qi Ng"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Getting Started

# Load packages

```{r cars}
library(tidyverse)
library(openintro)

```
# Exercise 1

Looking at these 3 histograms, the 2nd histogram graph shows more distrubution as the binwidth is smaller compare to the 3rd histogram, where there is only 4 bars showing various counts of flights with different departure delay numbers. Clearer separation of flights on what timeline is seen here.  The 1st histogram seems to look like a middleground or default. The huge majority of the departure delays are close to zero which tells me that most flight leave on time or within a few minutes of their schedule with some few outliers. shows the true form where there are 7 bars that indicate the number of flights for each one that are within a certain range of dep_delay numbers.I would say the 3rd one is a more aggregated view and looke like it is grouping a large number of flights into one bin, making the data less disperse overall and harder to see the close detail of the time of delay departures. 
